---
title: "STA 207"
author: "Romy Gou, Seoyoung Kim, Zichen Liu, Mario Arce Acosta"
date: "`r Sys.Date()`"
output: rmdformats::html_clean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
#install.packages("rmdformats")
```

## Background

```{r}

```

Describe the source of data Describe the experiment including potential caveats

## Question of Interest

Clearly state the question(s) of interest you aim to answer. Explain why they are of interest.

## Descriptive analysis

Provide informative summary statistics and visualizations. Use these summaries to motivate or justify subsequent choices in data preprocessing and modeling

## Inferential analysis

Propose and justify statistical models suitable for your questions. Clearly state model assumptions. Conduct formal inference and interpret results in context.

## Sensitivity analysis

Investigate plausibility of model assumptions Discuss robustness of results to alternative choices of models or assumptions

## Conclusion

Summarize findings.
