---
title: "STA 207"
author: "Romy Gou, Seoyoung Kim, Zichen Liu, Mario Arce Acosta"
date: "`r Sys.Date()`"
output: rmdformats::html_clean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
#install.packages("rmdformats")
```

## Background

```{r}

```

Describe the source of data Describe the experiment including potential caveats

## Question of Interest

Clearly state the question(s) of interest you aim to answer. Explain why they are of interest.

## Descriptive analysis

Provide informative summary statistics and visualizations. Use these summaries to motivate or justify subsequent choices in data preprocessing and modeling

## Inferential analysis

Propose and justify statistical models suitable for your questions. Clearly state model assumptions. Conduct formal inference and interpret results in context.

### Hierarchical Linear Model
$$
\begin{equation}
\begin{aligned}
    Y_{ijk} &= \big[ \gamma_{000} + \gamma_{001}(\text{SmallClass}_k) + \gamma_{010}(\text{Minority}_{jk}) + \gamma_{011}(\text{SmallClass}_k \times \text{Minority}_{jk}) + \gamma_{020}(\text{FreeLunch}_{jk}) \big] \\
    &+ \big[ \gamma_{100} + \gamma_{101}(\text{SmallClass}_k) + \gamma_{110}(\text{Minority}_{jk}) + \gamma_{111}(\text{SmallClass}_k \times \text{Minority}_{jk}) \big] \text{Grade2}_{ijk} \\
    &+ \big[ \gamma_{200} + \gamma_{201}(\text{SmallClass}_k) + \gamma_{210}(\text{Minority}_{jk}) + \gamma_{211}(\text{SmallClass}_k \times \text{Minority}_{jk}) \big] \text{Grade3}_{ijk} \\
    &+ v_{00k} + u_{0jk} + \epsilon_{ijk}
\end{aligned}
\end{equation}
$$

## Sensitivity analysis

Investigate plausibility of model assumptions Discuss robustness of results to alternative choices of models or assumptions

## Conclusion

Summarize findings.
