---
title: "STA 207"
author: "Romy Gou, Seoyoung Kim, Zichen Liu, Mario Arce Acosta"
date: "`r Sys.Date()`"
output: rmdformats::html_clean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
#library(MASS)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(patchwork)
library(knitr)
library(kableExtra)
library(car)
library(ggalluvial)
library(haven)
library(lme4)
library(lmerTest)
# install.packages("performance")
library(performance)
library(webshot2)
library(modelsummary)
library(broom.mixed)
library(stringr)
```

## Background

Describe the source of data Describe the experiment including potential caveats

## Question of Interest

Clearly state the question(s) of interest you aim to answer. Explain why they are of interest.

## Data Cleaning and Preparing

```{r Read Data}
STAR_Students <- read_sav("STAR_Students.sav")     
STAR_Students

STAR_K_3_Schools <- read_sav("STAR_K-3_Schools.sav")
STAR_K_3_Schools
```

```{r Restrict to three grades}
star_first_to_third_grade <- STAR_Students %>% 
  select(1:7, starts_with("g1"), starts_with("g2"), starts_with("g3")) %>%
  na.omit(race, g1treadss, g1tmathss, g1tlistss, g2treadss, g2tmathss, g2tlistss, g3treadss, g3tmathss, g3tlistss)
```

```{r Cleaning and }
STAR_long <- star_first_to_third_grade %>%
  pivot_longer(
    cols = matches("^g\\d"), 
    names_to = c("grade", ".value"), 
    names_pattern = "g(\\d)(.*)"
  ) %>%
  mutate(
    small_class = as.factor(ifelse(classtype==1, 1, 0)),
    minority = as.factor(ifelse(race!=1, 1, 0)),
    classtype = as.factor(classtype),
    grade = as.factor(grade),
    schid = as.factor(schid),
    stdntid = as.factor(stdntid)
  ) %>%
  group_by(stdntid) %>%
  filter(n_distinct(schid, na.rm = TRUE) == 1) %>%
  ungroup()

# To see how many students you have left:
print(paste("Students remaining:", n_distinct(STAR_long$stdntid)))
```

## Descriptive analysis

Even though STAR intended random assignment, many students did not stay in their assigned class type.

### Noncompliance and Attrition

```{r Alluvial Plot}
#| fig-width: 8
#| fig-height: 5
# drop g1classtype NA
sum(is.na(STAR_Students$g1classtype))

data_alluvium <- subset(STAR_Students, select = c(gkclasstype, g1classtype, g2classtype, g3classtype))

class_levels <- c("small", "regular", "regular-aide")

data_alluvium <- data_alluvium %>%
    mutate(across(everything(),
                  ~factor(as.numeric(.), levels = c(1, 2, 3),
                          labels = class_levels))) %>%
    mutate(across(everything(),
                  ~fct_explicit_na(., na_level = "Unknown"))) %>%
    # Remove any row where at least one grade is Unknown
    filter(if_all(everything(), ~ . != "Unknown"))


class_levels <- c("small", "regular", "regular-aide")
data_alluvium <- data_alluvium %>%
    mutate(across(everything(), 
                  ~factor(as.numeric(.), levels = c(1, 2, 3), 
                          labels = class_levels))) %>%
    mutate(across(everything(), 
                  ~fct_explicit_na(., na_level = "Unknown")))

data_summary <- data_alluvium %>%
    count(gkclasstype, g1classtype, g2classtype, g3classtype, name = "Freq") %>%
    mutate(path_id = row_number())  

data_long <- data_summary %>%
    pivot_longer(cols = starts_with("g"), names_to = "grade", values_to = "class")

data_long$grade <- factor(data_long$grade,
            levels = c("gkclasstype", "g1classtype", "g2classtype","g3classtype"),
            labels = c("Kindergarten", "Grade 1", "Grade 2", "Grade 3"))

ggplot(data_long, aes(x = grade, stratum = class, alluvium = path_id, y = Freq, fill = class)) +
    geom_flow(stat = "alluvium", alpha = 0.7) + 
    geom_stratum() + 
    scale_fill_manual(values = c("gold2", "skyblue2", "darkseagreen3")) +
    theme_minimal() +
    labs(title = "Fig3.1: Alluvial Plot of Students' Transfer", x = "Grade", y = "") +
    theme(legend.position = "right",
          axis.text.x = element_text(hjust = 0.5),
          axis.text.y = element_blank(),  
          axis.ticks.y = element_blank(), 
          axis.ticks.x = element_blank(), 
          panel.grid.major.y = element_blank(),
          panel.grid.minor.y = element_blank(),
          panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank(),
          plot.title = element_text(hjust = 0.5))
```

Drop rows with STAR_Students\$g1classtype=NA (4772 rows dropped)

Originally, STAR planned to keep students in the same classroom type from kindergarten through third grade, but that didn’t last. After parents raised concerns, noted in Hanushek (1999), the study was modified so that at the start of first grade, all students in regular‑size classes—whether or not they had an aide—were randomly reassigned to either a regular class with an aide or one without. Students who began in small classes weren’t part of this reshuffling and stayed in small classes for the rest of the study.

Beyond this official change, schools didn’t always follow the assignment rules in practice. Some, especially smaller ones, made informal adjustments once the year began. These weren’t parent-driven; they were practical decisions by administrators or teachers trying to manage real‑world constraints. Schools moved students to balance class sizes, address classroom dynamics, and etc. None of this was part of the experimental protocol, but it still mattered because it introduced nonrandom noncompliance. Finn & Achilles (1999) note that this weakens the clean randomization and can bias causal estimates.

### Class Size vs Class Type Density

```{r}
#| fig-width: 8
#| fig-height: 5

STAR_viz <- STAR_long %>%
  mutate(
    # Swapped 'small_class' for 'classtype' to capture all three assignments
    assigned_type = factor(classtype, 
                           levels = c(1, 2, 3), # Standard STAR numeric codes
                           labels = c("Small Class", "Regular Class", "Regular + Aide")),
    grade_label = paste("Grade", grade) 
  )

# 2. Create the hybrid Violin + Boxplot
class_density <- ggplot(STAR_viz, aes(x = assigned_type, y = classsize, fill = assigned_type)) +
  
  # Violin Plot
  geom_violin(alpha = 0.4, color = NA, trim = FALSE) + 
  
  # Boxplot
  # width = 0.2 makes it thin so it fits perfectly inside the "belly" of the violin
  geom_boxplot(width = 0.2, alpha = 0.8, color = "black") + 
  
  # Facet by grade
  facet_wrap(~ grade_label) +
  
  # Formatting & Aesthetics
  theme_minimal(base_size = 14) + 
  labs(title = "Manipulation Check: Actual Class Size Distribution",
       subtitle = "STAR Project Compliance Check by Grade",
       x = "Assigned Class Type",
       y = "Actual Number of Students in Class") +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank()) + 
  
  # The Colorful Part: Added a third color ("#C77CFF" / Purple) for the Aide class
  scale_fill_manual(values = c("Small Class" = "#F8766D", 
                               "Regular Class" = "#00BFC4", 
                               "Regular + Aide" = "#C77CFF")) +
  
  # Keeping this here in case you add geom_jitter() back into your plot later!
  scale_color_manual(values = c("Small Class" = "#D95F56", 
                                "Regular Class" = "#008B8E", 
                                "Regular + Aide" = "#A566D6"))

# Add this to the very bottom of your ggplot code:
class_density <- class_density +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

class_density

ggsave(
  filename = "class_density.png", 
  plot     = class_density
  )
```

*Provide informative summary statistics and visualizations. Use these summaries to motivate or justify subsequent choices in data preprocessing and modeling*

## Inferential analysis

### Hierarchical Linear Model

$$ \begin{equation} \begin{aligned}     Y_{ijk} &= \big[ \gamma_{000} + \gamma_{001}(\text{SmallClass}_k) + \gamma_{010}(\text{Minority}_{jk}) + \gamma_{011}(\text{SmallClass}_k \times \text{Minority}_{jk}) + \gamma_{020}(\text{FreeLunch}_{jk}) \big] \\     &+ \big[ \gamma_{100} + \gamma_{101}(\text{SmallClass}_k) + \gamma_{110}(\text{Minority}_{jk}) + \gamma_{111}(\text{SmallClass}_k \times \text{Minority}_{jk}) \big] \text{Grade2}_{ijk} \\     &+ \big[ \gamma_{200} + \gamma_{201}(\text{SmallClass}_k) + \gamma_{210}(\text{Minority}_{jk}) + \gamma_{211}(\text{SmallClass}_k \times \text{Minority}_{jk}) \big] \text{Grade3}_{ijk} \\     &+ v_{00k} + u_{0jk} + \epsilon_{ijk} \end{aligned} \end{equation} $$ where $i = 0, 1, 2$ and represents the different grade levels. $i=0$ is the 1st grade and our baseline for this longitudinal analysis.

```{r HLM Specification}

# Model 1: Base (Categorical)
STAR_long$classtype = relevel(STAR_long$classtype, ref=2)

hlm_categorical <- lmer(
  tmathss ~ freelunch + classtype * minority * grade + (1 | schid) + (1 | stdntid), 
  
  data = STAR_long,
  REML = TRUE
)
summary(hlm_categorical)

# Model 2: Dummy
hlm_dummy <- lmer(
  tmathss ~ small_class * minority * grade + 
  (1 | schid) + (1 | stdntid), 
  data = STAR_long,
  REML = TRUE
)
summary(hlm_dummy)

# Model 3: Dummy + Free/Reduced Lunch
hlm_dummy_2 <- lmer(
  tmathss ~ freelunch + small_class * minority * grade + 
  (1 | schid) + (1 | stdntid), 
  data = STAR_long,
  REML = TRUE 
)
summary(hlm_dummy_2)

# Model 4: Null Model
hlm_null <- lmer(tmathss ~ 1 + (1 | schid) + (1 | stdntid), data = STAR_long, REML = TRUE)
summary(hlm_null)
```

```{r}
#| fig-width: 8
#| fig-height: 5

# =========================================================
# LEVEL 1 PLOTS
# =========================================================
# 1A. Homoscedasticity (ggplot version)
homo_df <- data.frame(Fitted = fitted(hlm_dummy_2), Residuals = resid(hlm_dummy_2))
p1_homo <- ggplot(homo_df, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.2, color = "black") +
  geom_smooth(method = "loess", color = "blue", se = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "L1: Homoscedasticity", x = "Fitted Values", y = "Residuals")

# 1B. ACF Plot (ggplot version)
acf_calc <- acf(resid(hlm_dummy_2), plot = FALSE)
acf_df <- data.frame(lag = acf_calc$lag, acf = acf_calc$acf)
ci_bound <- qnorm((1 + 0.95)/2) / sqrt(acf_calc$n.used)

p2_acf <- ggplot(acf_df, aes(x = lag, y = acf)) +
  geom_segment(aes(xend = lag, yend = 0)) +
  geom_hline(yintercept = c(ci_bound, -ci_bound), color = "blue", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "black") +
  theme_minimal() +
  labs(title = "L1: Autocorrelation (ACF)", x = "Lag", y = "ACF")

# 1C. Normality of Residuals
res_df <- data.frame(residuals = resid(hlm_dummy_2))
p3_norm1 <- ggplot(res_df, aes(x = residuals)) +
  geom_density(fill = "steelblue", alpha = 0.5, color = NA) +
  stat_function(fun = dnorm, 
                args = list(mean = mean(res_df$residuals), sd = sd(res_df$residuals)), 
                color = "red", linewidth = 1, linetype = "dashed") +
  theme_minimal() +
  labs(title = "L1: Normality of Residuals", x = "Residuals", y = "Density")

# =========================================================
# LEVEL 2 & 3 PLOTS
# =========================================================

# Extract random intercepts
student_intercepts <- ranef(hlm_dummy_2)$stdntid
school_intercepts <- ranef(hlm_dummy_2)$schid

# 2. Level 2 (Student) Normality
std_df <- data.frame(intercepts = student_intercepts$`(Intercept)`)
p4_norm2 <- ggplot(std_df, aes(x = intercepts)) +
  geom_density(fill = "seagreen", alpha = 0.5, color = NA) +
  stat_function(fun = dnorm, 
                args = list(mean = mean(std_df$intercepts, na.rm=TRUE), sd = sd(std_df$intercepts, na.rm=TRUE)), 
                color = "red", linewidth = 1, linetype = "dashed") +
  theme_minimal() +
  labs(title = "L2: Student Intercepts", x = "Student Intercepts", y = "Density")

# 3. Level 3 (School) Normality
sch_df <- data.frame(intercepts = school_intercepts$`(Intercept)`)
p5_norm3 <- ggplot(sch_df, aes(x = intercepts)) +
  geom_density(fill = "darkorange", alpha = 0.5, color = NA) +
  stat_function(fun = dnorm, 
                args = list(mean = mean(sch_df$intercepts), sd = sd(sch_df$intercepts)), 
                color = "red", linewidth = 1, linetype = "dashed") +
  theme_minimal() +
  labs(title = "L3: School Intercepts", x = "School Intercepts", y = "Density")

# Top row (1x2): Homoscedasticity | ACF
# Bottom row (1x3): L1 Normality | L2 Normality | L3 Normality

final_diagnostic_grid <- (p1_homo | p2_acf) / (p3_norm1 | p4_norm2 | p5_norm3)

# Add the universal title
final_diagnostic_grid <- final_diagnostic_grid + plot_annotation(
  title = "Hierarchical Linear Model Diagnostics",
  subtitle = "Level 1 (Observations), Level 2 (Students), and Level 3 (Schools)"
)

# Display the grid in your R viewer
final_diagnostic_grid
```

```{r}
icc(hlm_null, by_group = TRUE)
```

```{r}
# Define your models in a list
hlm_models <- list(
  "HLM: Categorical Class Type" = hlm_categorical,
  "HLM: Binary Class Type" = hlm_dummy,
  "HLM: Binary Class Type + Free Lunch Indicator" = hlm_dummy_2
  )
```

\

```{r}
#| fig-width: 8
#| fig-height: 5
#| 
# ==========================================
# 1. PREPARE THE DATA
# ==========================================
coef_data <- tidy(hlm_dummy_2, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)")

coef_clean <- coef_data %>%
  mutate(
    Effect_Group = ifelse(str_detect(term, ":"), "Interaction Effects", "Main Effects"),
    Clean_Term = case_when(
      term == "freelunch" ~ "Free/Reduced Lunch", 
      term == "small_class1" ~ "Small Class",
      term == "minority1" ~ "Minority Student",
      term == "grade2" ~ "Grade 2",
      term == "grade3" ~ "Grade 3",
      term == "small_class1:minority1" ~ "Small Class × Minority",
      term == "small_class1:grade2" ~ "Small Class × Grade 2",
      term == "small_class1:grade3" ~ "Small Class × Grade 3",
      term == "minority1:grade2" ~ "Minority × Grade 2",
      term == "minority1:grade3" ~ "Minority × Grade 3",
      term == "small_class1:minority1:grade2" ~ "Small Class × Minority × Grade 2",
      term == "small_class1:minority1:grade3" ~ "Small Class × Minority × Grade 3",
      TRUE ~ term
    )
  ) %>%
  arrange(Effect_Group, estimate) %>%
  mutate(Clean_Term = factor(Clean_Term, levels = unique(Clean_Term)))

# ==========================================
# MAIN EFFECTS ONLY
# ==========================================
main_data <- coef_clean %>% filter(Effect_Group == "Main Effects")

plot_main <- ggplot(main_data, aes(x = estimate, y = Clean_Term)) +
  # Reduced linewidth
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.8) +
  # Reduced size and linewidth
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), color = "#2C3E50", size = 0.5, linewidth = 0.8) +
  
  # Reduced text size from 6 to 3.5
  geom_text(aes(label = sprintf("%.2f", estimate)), 
            vjust = -1.2, size = 3.5, fontface = "bold", color = "#2C3E50") +
  
  # Reduced base_size from 18 to 12
  theme_minimal(base_size = 12) + 
  theme(
    panel.grid.minor = element_blank(),
    # Reduced axis text size from 16 to 10
    axis.text.y = element_text(face = "bold", size = 10)
  ) +
  labs(
    title = "Main Effects: Baseline Predictors of Math Achievement",
    x = "Point Estimate (Impact on Math Score)",
    y = NULL
  ) +
  scale_y_discrete(expand = expansion(mult = c(0.1, 0.15)))
plot_main

# ==========================================
# INTERACTION EFFECTS ONLY
# ==========================================
interaction_data <- coef_clean %>% filter(Effect_Group == "Interaction Effects")

plot_interactions <- ggplot(interaction_data, aes(x = estimate, y = Clean_Term)) +
  # Reduced linewidth
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.8) +
  # Reduced size and linewidth
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), color = "#E74C3C", size = 0.5, linewidth = 0.8) +
  
  # Reduced text size from 6 to 3.5
  geom_text(aes(label = sprintf("%.2f", estimate)), 
            vjust = -1.2, size = 3.5, fontface = "bold", color = "#E74C3C") +
  
  # Reduced base_size from 18 to 12
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    # Reduced axis text size from 16 to 10
    axis.text.y = element_text(face = "bold", size = 10)
  ) +
  labs(
    title = "Interaction Effects: Compounding Variables Over Time",
    x = "Point Estimate (Impact on Math Score)",
    y = NULL
  ) +
  scale_y_discrete(expand = expansion(mult = c(0.05, 0.1)))
plot_interactions
```

Propose and justify statistical models suitable for your questions. Clearly state model assumptions. Conduct formal inference and interpret results in context.

## Sensitivity analysis

*Investigate plausibility of model assumptions Discuss robustness of results to alternative choices of models or assumptions*

## Conclusion

Summarize findings.
